The Effects of Stellar Binarity on Transit Survey Occurrence Rates

### 2017-09-25 paper/ms.tex
include corrected plot; update text

\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/q_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the mass ratio for a magnitude limited 
		sample of binary stars. The underlying mass ratios are drawn from 
		Eq.~\ref{eq:mass_ratio} -- a \textit{uniform} distribution in a 
		volume-limited sample.
		The entire bias can be understood analytically 
		(Eq.~\ref{eq:pdf_q_mag_limited}) to good precision. 
        The numerical comparison uses the empirical mass-luminosity relation 
        shown in Fig.~\ref{fig:mass_luminosity}.
	}
	\label{fig:q_distribn_mag_limited}
\end{figure}

\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/gammaR_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the luminosity ratio for a magnitude limited 
		sample of binary stars. Despite the preference in the sample towards 
		large $q$ binaries (Fig.~\protect\ref{fig:q_distribn_mag_limited}), 
		the bias is towards small $\gamma_R$ binaries because of the steepness 
		of the mass-luminosity relation.
        The underlying mass ratios are uniform in volume.
        The numerical comparison uses the empirical mass-luminosity relation 
        shown in Fig.~\ref{fig:mass_luminosity}.       
    }
	\label{fig:gammaR_distribn_mag_limited}
\end{figure}

@@

Marginalizing the joint distribution over distance, and assuming $L\propto 
M^{3.5}$,
\begin{align}
{\rm prob}_{\rm ml}(q) &= 
\int_{0}^{d_{\rm max}(q)} {\rm prob}_{\rm ml}(r,q) \,{\rm d}r \\
&= \mathcal{Z}\cdot (1+q^{3.5})^{3/2}\quad {\rm if\ }0.1<q\leq 1,
\label{eq:pdf_q_mag_limited}
\end{align}
and otherwise 0, for $\mathcal{Z}$ the appropriate normalization constant.
This marginalized distribution~--~the distribution function of mass ratios for 
binaries in a magnitude-limited sample~--~is plotted in 
Fig.~\ref{fig:q_distribn_mag_limited}.

@@

Doing some algebra, and assuming $\gamma_R = q^\alpha$, this can be shown to be
\begin{align}
{\rm prob}(N_d) = \frac{2\mathcal{Z}}{3\alpha}& N_d^{1/3} \left(\frac{3}{4\pi 
	n_d}\right)^{4/3} d_{\rm max,s}^{-4} \ \nonumber\\
&\times \left[ 
	\left(\frac{3 N_d}{4\pi n_d}\right)^{2/3}  d_{\rm max,s}^{-2} - 1
 \right]^{\frac{1-\alpha}{\alpha} }
\end{align}
over the interval
\begin{equation}
N_d^{\rm lower} = \frac{4 \pi n_d}{3} (\sqrt{0.1^3 + 1} d_{\rm max,s})^3,
\end{equation}
\begin{equation}
N_d^{\rm upper} = \frac{4 \pi n_d}{3} (\sqrt{2}d_{\rm max,s})^3,
\end{equation}
and outside the stated interval ${\rm prob}(N_d) = 0$.

An interesting intermediate result is the magnitude-limited luminosity ratio 
distribution,
\begin{equation}
{\rm prob_{ml}}(\gamma_R) = \frac{\mathcal{Z}}{\alpha} (1+\gamma_R) 
\gamma_R^{\frac{1-\alpha}{\alpha}},
\end{equation}
for $(0.1)^{\alpha}<\gamma_R<1$, and otherwise zero.
For $\alpha=3.5$, ${\rm prob_{ml}}(\gamma_R) \propto 
(1+\gamma_R)\gamma_R^{-5/7}$, which tells us the distribution will be highly 
biased towards low-light ratio binaries.
The distribution is plotted in Fig.~\ref{fig:gammaR_distribn_mag_limited}.

The reason we derived all this is that we wanted to know the number of 
searchable binary stars.
The expected value for the number of double star systems in the sample is
\begin{equation}
\langle N_d \rangle = \int_{0}^{\infty} N_d\, {\rm prob}(N_d) \,{\rm d}N_d.
\end{equation}
With some algebra, and letting $\alpha=3.5$, this can be expressed as
\begin{equation}
\langle N_d \rangle = \frac{2 \mathcal{Z}}{7 R^{1/2}}
\int_{(0.1^\alpha + 1)^{1/3}}^{\sqrt{2}} u^{3/2} (u-1)^{5/7} \,{\rm d}u ,
\label{eq:integral_Nd}
\end{equation}
for
\begin{equation}
R \equiv \left(\frac{3}{4\pi n_d}\right)^{2/3} d_{\rm max,s}^{-2}.
\end{equation}
The integral in Eq.~\ref{eq:integral_Nd} has a closed-form solution in terms 
of Gauss hypergeometric functions, $_2F_1$. Once evaluated, the integral takes 
a numerical value of $0.1829$.
The normalization of

%TODO GIVE THE EXPLICIT FORM, SEMI-ANALYTICALLY.

While a closed analytic expression for $\langle N_d \rangle$ does exist, it is 
messy, and it does not yield much intuition. Instead, summarizing the important 
points and supporting them numerically:
\begin{itemize}
	\item In a volume-limited sample of binary star systems in which the 
	primary mass is fixed, and the mass ratio is drawn from a bounded uniform 
	distribution, the distribution of $\gamma_R$ will be biased towards low 
	values ($\gamma_R \approx 0.1$). This is shown in 
	Fig.~\ref{fig:gammaR_distribn_vol_limited}.
	\item In a magnitude-limited sample of binary star systems in which the 
	primary mass is fixed, and the mass ratio of the \textit{population} is 
	drawn from a bounded uniform distribution, the observed distribution of 
	mass ratios will be biased towards high values. You will see more twins,
	because they are detectable out to a greater distance\footnote{This is a 
	scary systematic w.r.t. the claimed intrinsic excess of twin binaries.}. 
	This is shown in Fig.~\ref{fig:q_distribn_mag_limited}, and explained 
	analytically below.
	\item In a magnitude-limited sample of binary star systems in which the 
	primary mass is fixed, the distribution of $\gamma_R$ will be biased 
	towards low values ($\approx 0.1$), but less so than in a 
	volume-limited sample. This is shown in 
	Fig.~\ref{fig:gammaR_distribn_mag_limited}.
\end{itemize}

In passing, the bias in ${\rm prob}(\gamma_R)$ towards low luminosity ratios 
can be seen analytically. If we assume $\gamma_R = q^\alpha$, then
\begin{equation}
{\rm prob}(\gamma_R) = \frac{1}{9 \alpha} \ \gamma_R^{\frac{1-\alpha}{\alpha}}
\quad {\rm for\ } (0.1)^\alpha < \gamma_R < 1,
\end{equation}
and otherwise zero. For instance if $\alpha = 3$, ${\rm prob}(\gamma_R) 
\propto \gamma_R^{-2/3}$ and the domain extends from $1$ to 
$10^{-3}$, where the probability distribution peaks.

\paragraph{How many stars are in the sample?}

We return to the original question: how many stars?
\begin{equation}
\langle N_{\rm stars} \rangle = N_s + 2 \langle N_d \rangle,
\end{equation}
as before.
Since we are not giving analytic expressions for either of the $\langle \ldots 
\rangle$ terms, we just note that in the case of a population with a single 
$\gamma_R$ value, $N_d/N_s = {\rm BF} \times (1 + \gamma_R)^{3/2}$. A lower
bound for this ratio will always be the binary fraction.
For a single run of a Monte Carlo code, where ${\rm mean}(\gamma_R) = 0.28$, 
${\rm median}(\gamma_R) = 0.13$, and the distribution was that given in 
Fig.~\ref{fig:gammaR_distribn_mag_limited}, the observed ratio was $N_d/N_s = 
0.59$.
This corresponds to a single-valued population with $\gamma_R \sim 0.20$, a
number between the mean and median of the true distribution.
It means $\sim 1.2$ stars in binary star systems for every star in a single 
star system from this sample.

We also note in passing that $N_d/N_s = 0.59$ is a smaller ratio than for the 
case of a fixed $\gamma_R$ population with $\gamma_R = 1$, which gave $N_d/N_s 
= 1.27$. The difference is that in the latter population, the volume of 
searchable stars is greater. Based on the fixed-$\gamma_R$ population scaling 
law $N_d / N_s \propto (1+\gamma_R)^{3/2}$, we 
would expect the ratio of single stars to be $\sim (2/1.1)^{3/2} = 2.4$ 
times smaller, which is roughly (though not exactly) what we observe.

@@

By default, Galaxia does not include binaries.
However, from Fig.~\ref{fig:rmag_distribn}, we've seen that its number counts 
as a function of apparent magnitude are good.
So we assign binarity to the Galaxia stars as follows.
\begin{itemize}
	\item Assume a binary fraction ${\rm BF} = 0.45$ [Raghavan et al 2010].
	\item Ignore higher order multiples.
	\item If a star is drawn to be a binary:
	\subitem Its reported magnitude becomes a system magnitude.
	\subitem Draw the binary's mass ratio $q$ from the probability distribution 
	function for a magnitude limited survey, i.e. from 
	Fig.~\ref{fig:q_distribn_mag_limited}. This is sensible so long as the pdf 
	is independent of the primary's mass. We assume this to be the case over 
	$0.7-1.3M_\odot$\footnote{It was not immediately self-evident to me that 
	this distribution was the correct choice, since for fixed mass ratio as the 
	primary mass changes so does the detectable volume. However, as long as 
	${\rm prob}(q)$ is independent of the primary's mass, the average 
	distribution over all the possible detectable volumes will remain the 
	same.}.
	\subitem For $q \geq 5/7$, it turns out the above is sufficient information 
	to analytically specify the individual stellar masses, and thus 
	luminosities with the relation shown in Fig.~\ref{fig:mass_luminosity}.
	\subitem For $q < 5/7$, the individual stellar masses are not uniquely 
	specified\footnote{See the derivation of 2017/08/25. This is because of the 
	mass luminosity relation we specified, and the fact that we have $q$, and 
	$L_d$, but want individual masses, and individual luminosities.}. Thus for 
	this latter fraction of the population, we draw the 
	primary mass from the pdf of single star primary masses.
	\subitem With the individual masses known, use Eq.~\ref{eq:mass_radius} to 
	assign a radius to each star in the binary system.
\end{itemize}

>>>>

\begin{figure}[!t]
	\begin{center}
		\includegraphics[width=0.5\textwidth]{figures/q_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the mass ratio for a magnitude limited 
		sample of binary stars. The underlying mass ratios are drawn from 
		Eq.~\ref{eq:mass_ratio} -- a \textit{uniform} distribution in a 
		volume-limited sample.
		The entire bias can be understood analytically 
		(Eq.~\ref{eq:pdf_q_mag_limited}) to good precision. 
        The numerical comparison uses the empirical mass-luminosity relation 
        shown in Fig.~\ref{fig:mass_luminosity}.
	}
	\label{fig:q_distribn_mag_limited}
\end{figure}

\begin{figure}[!t]
	\begin{center}
		\includegraphics[width=.5\textwidth]{figures/gammaR_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the luminosity ratio for a magnitude limited 
		sample of binary stars. Despite the preference in the sample towards 
		large $q$ binaries (Fig.~\protect\ref{fig:q_distribn_mag_limited}), 
		the bias is towards small $\gamma_R$ binaries because of the steepness 
		of the mass-luminosity relation.
        The underlying mass ratios are uniform in volume.
        The numerical comparison uses the empirical mass-luminosity relation 
        shown in Fig.~\ref{fig:mass_luminosity}.       
    }
	\label{fig:gammaR_distribn_mag_limited}
\end{figure}

@@

Marginalizing the joint distribution over distance, and assuming $L\propto 
M^{3.5}$,
\begin{align}
{\rm prob}_{\rm ml}(q) &= 
\int_{0}^{d_{\rm max}(q)} {\rm prob}_{\rm ml}(r,q) \,{\rm d}r \\
&= \mathcal{Z}\cdot (1+q^{7/2})^{3/2}\quad {\rm if\ }0.1<q\leq 1,
\label{eq:pdf_q_mag_limited}
\end{align}
and otherwise 0, for $\mathcal{Z}$ the appropriate normalization constant.
This marginalized distribution~--~the distribution function of mass ratios for 
binaries in a magnitude-limited sample~--~is plotted in 
Fig.~\ref{fig:q_distribn_mag_limited}.

@@

Doing some algebra, and assuming $\gamma_R = q^\alpha$, this can be shown to be
\begin{align}
{\rm prob}(N_d) \propto N_d^{2/3} &\left(\frac{3}{4\pi 
	n_d}\right)^{5/3} d_{\rm max,s}^{-4} \ \nonumber\\
&\times \left[ 
	\left(\frac{3 N_d}{4\pi n_d}\right)^{2/3}  d_{\rm max,s}^{-2} - 1
 \right]^{\frac{1-\alpha}{\alpha} }
\end{align}
over the interval
\begin{equation}
N_d^{\rm lower} = \frac{4 \pi n_d}{3} (\sqrt{0.1^3 + 1} d_{\rm max,s})^3,
\end{equation}
\begin{equation}
N_d^{\rm upper} = \frac{4 \pi n_d}{3} (\sqrt{2}d_{\rm max,s})^3,
\end{equation}
and outside the stated interval ${\rm prob}(N_d) = 0$.

An interesting intermediate result is the magnitude-limited luminosity ratio 
distribution,
\begin{equation}
{\rm prob_{ml}}(\gamma_R) \propto (1+\gamma_R)^{3/2} 
\gamma_R^{\frac{1-\alpha}{\alpha}},
\end{equation}
for $(0.1)^{\alpha}<\gamma_R<1$, and otherwise zero.
For $\alpha=3.5$, ${\rm prob_{ml}}(\gamma_R) \propto 
(1+\gamma_R)^{3/2}\gamma_R^{-5/7}$, which tells us the distribution will be 
highly biased towards low-light ratio binaries, as shown in 
Fig.~\ref{fig:gammaR_distribn_mag_limited}.

The reason we derived all this is that we wanted to know the number of 
searchable binary stars.
The expected value for the number of double star systems in the sample is
\begin{equation}
\langle N_d \rangle = \int_{0}^{\infty} N_d\, {\rm prob}(N_d) \,{\rm d}N_d.
\end{equation}
Letting $\alpha=3.5$, this can be expressed as
\begin{equation}
\langle N_d \rangle \propto \mathcal{R}^{-3/2}
\int_{0.1^{7/2} + 1}^{2} u^{3/2} (u-1)^{5/7} \,{\rm d}u ,
\label{eq:integral_Nd}
\end{equation}
where non-dimensional constants are not shown, and $\mathcal{R}$ is defined as
\begin{equation}
\mathcal{R} \equiv \left(\frac{3}{4\pi n_d}\right)^{2/3} d_{\rm max,s}^{-2}.
\end{equation}
The integral in Eq.~\ref{eq:integral_Nd} has a closed-form solution in terms 
of the Gauss hypergeometric function, $_2F_1$. Once evaluated, and after 
normalizing ${\rm prob}(N_d)$, to 4 significant digits it gives
\begin{equation}
\langle N_d \rangle = 1.601 \left(\frac{4\pi n_d}{3}\right) d_{\rm max,s}^3.
\end{equation}
This gives the useful expression
\begin{align}
\frac{\langle N_d \rangle}{N_s} &= 1.601\, \frac{n_d}{n_s} \\
&= 1.601 \left(\frac{{\rm BF}}{1-{\rm BF}}\right).
\end{align}
Taking the sun-like star binary fraction of ${\rm BF} = 0.44$, this gives
$\langle N_d \rangle / N_s = 1.258$.
Compared to the binary-twin thought experiment of Sec.~\ref{sec:model_1}, in 
which the SNR-limited sample had $\approx 2.8$ times more binary systems than 
single systems, there are now $\approx 1.3$ times as many.
This is sensible, because the mean light ratio has decreased; it was 
previously 1, it is now $\langle \gamma_R \rangle \approx 0.27$.

\paragraph{What is the true occurrence rate?}

Returning to the original question, the total number of stars in the sample is
\begin{equation}
\langle N_{\rm stars,tot} \rangle = N_s + 2 \langle N_d \rangle.
\end{equation}
For stars in binary systems, only a fraction of them are of the desired type.
Defining the ``desired type'' to be $0.7<M_\star/M_\odot<1.3$, this can be 
evaluated by integrating Eq.~\ref{eq:pdf_q_mag_limited} from $0.7M_\odot$ to 
$M_\odot$.
The result is that $47.5\%$ of secondaries are of the desired type.
Thus the expected number of stars of the desired type in the sample is
\begin{equation}
\langle N_{\rm stars} \rangle = N_s + 1.475 \langle N_d \rangle.
\end{equation}

The number of planets of the desired type is then
\begin{equation}
\langle N_{\rm planets} \rangle = 
\Gamma_{t,s}N_s + \Gamma_{t,d}\langle N_d\rangle + 
0.475\Gamma_{t,d}w_{d2} \langle N_d\rangle,
\end{equation}
and the true occurrence rate is
\begin{align}
\Gamma_t &= \frac{\langle N_{\rm planets} \rangle }{\langle N_{\rm stars} 
\rangle} \\
\Gamma_t &= 
\frac{
	\Gamma_{t,s}N_s + (1+0.475w_{d2})\Gamma_{t,d}\langle N_d\rangle
	}{
	N_s + 1.475 \langle N_d \rangle
	}.
\end{align}

% % % % % % % % % % % % % % % % % % % % % % % % % % % %
\subsection{What is the measured occurrence rate?}

The observer who has never heard of binary star systems wants to find the 
occurrence rate of planets with radius $R_p$ and period $P$, orbiting stars 
with mass $0.7<M_\star/M_\odot < 1.3$ (with corresponding stellar radii and 
luminosities).
They detected $N_{\rm det}$ transit signals that appear to be from planets of 
this desired type.
The occurrence rate they then find, {\it i.e.} the number of planets divided 
by number of ``stars'' (really, stellar systems) is
\begin{equation}
\Gamma_{\rm a } = \frac{N_{\rm det}}{N_s + N_d} \times \frac{1}{f_g}.
\end{equation}

Unlike in Sec.~\ref{sec:model_1}, where the observer saw signals of only two 
apparent radii (undiluted and diluted), they will now find signals with a 
spectrum of apparent radii.
The dilutions $\mathcal{D}$ in binary systems, given by Eq.~\ref{eq:dilution}, 
vary depending on which star the planet orbits, and depending on the system's 
luminosity ratio $\gamma_R$.
Moreover, the apparent planetary radii will be affected by the assumptions the 
observer makes about the apparently-single star's radius, $R_{\star,{\rm a}}$.
We posit that the observer assumes $R_{\star,{\rm a}} = R_1$; in other words 
they assume that the primary star hosts any signal they observe.
The apparent planetary radii, $R_{\rm p, a}$, are then
\begin{equation}
R_{p,{\rm a}} = R_p 
\frac{R_1}
{R_{\star,{\rm t}}}
\,\mathcal{D}^{1/2}.
\end{equation}

The total number of planet detections of the desired type is 
\begin{equation}
\langle N_{\rm det}\rangle = 
N_{{\rm det},s} 
+ 
\langle N_{{\rm det},d}\rangle.
\end{equation}
As in Sec.~\ref{sec:model_1},
\begin{equation}
N_{{\rm det},s} = N_s \Gamma_{t,s} f_{s,g} f_{s,c},
\label{eq:N_det_s_model2}
\end{equation}
where the product $N_s \Gamma_{t,s}$ is the number of planets in the single 
star systems of the sample, $f_{s,g}$ is the geometric transit probability, 
and $f_{s,c}$ is the fraction of these transiting planets that are observed 
with signal to noise greater than the minimum detection threshold (the 
completeness). 
For our SNR-limited, varying-$\gamma_R$ survey, $f_{s,g} = R_1/a\equiv f_g$, 
and $f_{s,c}=1$.

For planets detected in binary systems, only a fraction will be ``of the 
desired type'', since only a fraction of the stars are of the desired type.
This is the same factor $f_{\rm desired}$ that we have shown is 0.475. Thus
\begin{equation}
\langle N_{{\rm det},d} \rangle
= 
(1+0.475w_{d2}) \langle N_d\rangle \Gamma_{t,d} f_{d,g} f_{d,c}.
\label{eq:N_det_d_model2}
\end{equation}
Now $(1+0.475w_{d2}) \langle N_d\rangle \Gamma_{t,d}$ is the number of 
planets of the desired type in the double star systems of the sample.
The geometric transit probability $f_{d,g}$ and completeness $f_{d,c}$ require 
more careful evaluation.
First, separate them into their sub-components about primaries ``d1'' and 
secondaries ``d2'':
\begin{align}
\langle N_{{\rm det},d} \rangle
= 
\langle N_d\rangle \Gamma_{t,d} ( f_{d1,g} f_{d1,c} 
+
0.475w_{d2}  f_{d2,g} f_{d2,c}
)
\end{align}

The geometric transit probability $f_{d,g}=R_1/a$ 
for our twin-binary survey, and $f_{d,c}=1/8$ by the geometric argument 
presented in the initial discussion of limiting distances.

\subsection{Errors on derived occurrence rates for Model \#2}

@@

By default, Galaxia does not include binaries.
However, from Fig.~\ref{fig:rmag_distribn}, we've seen that its number counts 
as a function of apparent magnitude are good.
So we assign binarity to the Galaxia stars as follows.
\begin{itemize}
	\item Assume a binary fraction ${\rm BF} = 0.44$ [Raghavan et al 2010].
	\item Ignore higher order multiples.
	\item If a star is drawn to be a binary:
	\subitem Its reported magnitude becomes a system magnitude.
	\subitem Draw the binary's mass ratio $q$ from the probability distribution 
	function for a magnitude limited survey, i.e. from 
	Fig.~\ref{fig:q_distribn_mag_limited}. This is sensible so long as the pdf 
	is independent of the primary's mass. We assume this to be the case over 
	$0.7-1.3M_\odot$\footnote{It was not immediately self-evident to me that 
	this distribution was the correct choice, since for fixed mass ratio as the 
	primary mass changes so does the detectable volume. However, as long as 
	${\rm prob}(q)$ is independent of the primary's mass, the average 
	distribution over all the possible detectable volumes will remain the 
	same.}.
	\subitem For $q \geq 5/7$, it turns out the above is sufficient information 
	to analytically specify the individual stellar masses, and thus 
	luminosities with the relation shown in Fig.~\ref{fig:mass_luminosity}.
	\subitem For $q < 5/7$, the individual stellar masses are not uniquely 
	specified\footnote{See the derivation of 2017/08/25. This is because of the 
	mass luminosity relation we specified, and the fact that we have $q$, and 
	$L_d$, but want individual masses, and individual luminosities.}. Thus for 
	this latter fraction of the population, we draw the 
	primary mass from the pdf of single star primary masses.
	\subitem With the individual masses known, use Eq.~\ref{eq:mass_radius} to 
	assign a radius to each star in the binary system.
\end{itemize}
